An Introduction to Return Oriented Programming

### 2023-02-17 paper.tex
changes

\section{Introduction}

%%%% 7. PAPER CONTENT %%%%
\section{Introduction}
Return Oriented Programming, abbreviated ROP is a type of buffer overflow attack that has been published in 2007 by Hovav Shacham.~\cite{ropfirstpaper} and has become a widely known buffer overflow technique since. It has been developed to circumvent the NX-BIT protection that protects the stack from being executed. The general consensus is that modern binaries are practically not vulnerable to buffer overflow attacks, but there is a lot of research surrounding breaking of these security measures that shows practical strength of these security measures does not equal the theoretical strength due to side channels, bugs or other exploits.~\cite{aslr} With high enough reward ROP may be a devastating technique for black-hats, because of that it is important to raise awareness about binary exploitation generally and ROP. Because of that this paper will demonstrate the underlying theory and demonstrate it with an attack on a vulnerable binary.

@@

These are only 10 Lines out of the 8244 lines found by the tool though and i purposefully filtered out some good and bad ones for demonstration. It is clearly visible that many candidates for ROP can be found, even in a file with a relatively small size of 72 kB. Though most of these gadgets are not all that useful because they often modify a lot of registers, possibly messing up the desired state. In most cases we can find suitable candidates using regular expressions, this will be demonstrated later in this section~\cref{subsec:filtering}.

@@

\Verb+int+ stands for interrupt, the interrupt \bltInlineVerb{int 0x80} causes a system call to be executed. System calls are kernelspace programs/operations that require higher privileges than what is available in a userspace program. Examples for system calls include io and \Verb+execve+ which allows to execute arbitary programs. In combination with \Verb+pop+, \Verb+mov+ and other instructions we can specify the concrete system call.~\cite{ropsla} One of the most powerful system calls for blackhats is bash since it allows permanently implementing malware or gain insight into files, it can be called with the argument \Verb+/bin/sh+. This will be demonstrated in~\cref{sec:attack}.

@@

In order to find the required gadgets we can use the tools directly or we can use regular expressions. In order to make this paper more general and easy to replicate i will be using regular expressions to find the desired gadgets.

@@

The following table describes what regex we can use to find the gadgets required for the attack.

@@

\section{Theory}
\subsection{Stack}
The following graphic~\cref{fig:stack} is an illustration of how the stack changes when injecting the payload. The buffer first has to be filled. In binary exploitation the letter \Verb+A+ is used for that most of the time, it has an easy to identify hexadecimal value of \bltInlineVerb{0x41}. It is important to note that without any special compiler options the stack will be aligned in \bltInlineVerb{dword}'s/16 Byte blocks. because of that the buffer has to be filled with more bytes than the buffer holds if $s \mod 16 \neq 0$ holds true, s being the buffer size in Bytes.
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.79\textwidth]{stackropoffsec.png}
  \caption{The stack when injecting the payload}
  \label{fig:stack}
\end{figure}
\\
\subsection{ROP Runtime Behaviour}
The following graphic~\cref{fig:executionatruntime} illustrates how the gadgets get executed once the instruction pointer \bltInlineVerb{eip} points to the \bltInlineVerb{ret} in \bltInlineVerb{main}. Once this happens the execution gets redirected to the first gadget and executes the instructions in it. As soon as \bltInlineVerb{eip} points to the \bltInlineVerb{ret} in the 1st gadget the address of the 2nd gadget is \bltInlineVerb{pop}'d into \bltInlineVerb{eip} and execution continues there, from there the same thing happens again until execution reaches the end of the last gadget.
\begin{figure}[h!]
  \centering
  \includegraphics[width=0.95\textwidth]{Ropchaineffect.png}
  \caption{The stack when injecting the payload}
  \label{fig:executionatruntime}
\end{figure}

@@

After specifying the goal and possibly simplifying it we have to determine the required program state. For the example in this paper we want to open a shell, for that the simplest way is to execute an \bltInlineVerb{execve} system call. The following program state~\cref{fig:stateforint} has to be achieved so the interrupt \bltInlineVerb{int 0x80} causes a shell to be opened.~\cite{pixis}~\cite{proggen-rop}

@@

From all the previous steps the payload got constructed with python~\cref{code:payloadbinsh}. As seen in the example we can define all the addresses, gadgets and other parameters as variables and reuse them in the \bltInlineVerb{pack} calls, this way changing a gadget only requires one value to be changed. The different

@@

The only protections that had to be disabled were stack canaries and ASLR. It is likely that there are systems still in use today which are vulnerable to this kind of attack due to not having these protections or the protections themselves being attackable. Since it allows almost arbitrary code execution it is very important to identify these devices and patch or replace them.

@@

The information wether ROP works with ASLR enabled is inconsistent. While trying this attack with \bltInlineVerb{/proc/sys/kernel/randomize_va_space} set to \bltInlineVerb{2} meaning full randomization the attack still seemed to work. The inconsistent information probably arises due to different approaches being used. With executables that have PIE enabled ROP is still possible but only with ASLR disabled~\cite{el-sherei}. With the compiler options used for this example PIE is disabled and ASLR seems to have no effect on the exploit. This is because the ASLR settings \Verb+1+ and \Verb+2+ only randomize shared libraries and PIE binaries~\cite{rny}, since the program has been compiled with the \bltInlineVerb{-static} option, which implicitly compiles the program to not be position independent, ASLR is not being used, even when activated.

\section{Protection}
\paragraph{Stack canaries}
Stack canaries are one of the most effective approaches against ROP, they are enabled by default and prevent most forms of buffer overflows, however, stack canaries can be based on a small entropy pool and can therfore be bruteforced with an effort significantly smaller than regular bruteforcing. Depending on the target it can still be profitable and possible to bruteforce it even with a big entropy pool and high randomness.
\paragraph{NX}
The activation of the NX bit has no effect on ROP since the program never executes code outside the segments marked with the \bltInlineVerb{CODE} flag like in a classical stack overflow attack.~\cite{ropsla}
\paragraph{ASLR}
According to a paper by Hovav Shacham et al. ASLR is a good protection against ROP in 64 bit binaries assuming no side channel leakage since 40 bit are available for randomizations of the libraries and code locations, however, 32 Bit binaries only use 16 Bit for randomization. Because of that they were able to perform a buffer overflow attack like ret2libc on an Apache server with an average of 216 seconds.~\cite{aslr}
%%%% 8. BILBIOGRAPHY %%%%
\bibliographystyle{alpha}
\bibliography{refs}
%%%% NOTES
% - Download abbrev3.bib and crypto.bib from https://cryptobib.di.ens.fr/
% - Use bilbio.bib for additional references not in the cryptobib database.
%   If possible, take them from DBLP.
\end{document}

>>>>

%%%% 7. PAPER CONTENT %%%%
\section{Introduction}
Return Oriented Programming, abbreviated ROP is a type of buffer overflow attack that has been published in 2007 by Hovav Shacham.~\cite{ropfirstpaper} and has become a widely known buffer overflow technique since. It has been developed to circumvent the NX-BIT protection that protects the stack from being executed. The general consensus is that modern binaries are practically not vulnerable to buffer overflow attacks, but there is a lot of research surrounding breaking of these security measures that shows practical strength of these security measures does not equal the theoretical strength due to side channels, bugs or other exploits.~\cite{aslr} Because of its power it is important to raise awareness about binary exploitation generally and ROP. Because of that this paper will explain the underlying theory and demonstrate it with an attack on a vulnerable binary.

@@

These are only 7 gadgets out of the 8180 gadgets found by the tool though and have been purposefully picked for demonstration. It is clearly visible that many candidates for ROP can be found, even in a file with a relatively small size of 72 kB. Though most of these gadgets are not all that useful because they often modify a lot of registers, possibly messing up the desired state. In most cases we can find suitable candidates using regular expressions, this will be demonstrated later in this section~\cref{subsec:filtering}.

@@

\Verb+int+ stands for interrupt, the interrupt \bltInlineVerb{int 0x80} causes a system call to be executed. System calls are kernelspace programs/operations that require higher privileges than what is available in a userspace program. Examples for system calls include io and \Verb+execve+ which allows to execute arbitary programs. In combination with \Verb+pop+, \Verb+mov+ and other instructions we can specify the concrete system call.~\cite{ropsla} One of the most powerful system calls for blackhats is execve with \bltInlineVerb{/bin/sh} as argument since it gives the attacker full system access once privileges have been elevated.

@@

In order to find the required gadgets we can use the tools directly or we can use regular expressions. In order to make this paper more general and easy to replicate the method using regular expressions will be demonstrated.

@@

The following examples demonstrate how regular expressions can be used to search gadgets.

@@

\section{Theory}
\subsection{Stack}
Graphic~\cref{fig:stack} is an illustration of how the stack changes when injecting the payload. The buffer first has to be filled. In binary exploitation the letter \Verb+A+ is used for that most of the time, it has an easy to identify hexadecimal value of \bltInlineVerb{0x41}. It is important to note that the padding required may be bigger than the buffer due to compiler based stack alignment.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{stackropoffsec.png}
  \caption{The stack when injecting the payload}
  \label{fig:stack}
\end{figure}
\subsection{ROP Runtime Behaviour}
Graphic~\cref{fig:executionatruntime} illustrates how the gadgets get executed once the instruction pointer \bltInlineVerb{eip} points to the \bltInlineVerb{ret} in \bltInlineVerb{main}. Once this happens the execution gets redirected to the first gadget and executes the instructions in it. As soon as \bltInlineVerb{eip} points to the \bltInlineVerb{ret} in the 1st gadget the address of the 2nd gadget is \bltInlineVerb{pop}'d into \bltInlineVerb{eip} and execution continues there, from there the same thing happens again until execution reaches the end of the last gadget.
\begin{figure}[H]
  \centering
  \includegraphics[width=0.95\textwidth]{Ropchaineffect.png}
  \caption{The stack when injecting the payload}
  \label{fig:executionatruntime}
\end{figure}

@@

After specifying the goal and possibly simplifying it we have to determine the required program state. For the example in this paper we want to open a shell, for that the simplest way is to execute an \bltInlineVerb{execve} system call. The program state illustrated in graphic~\cref{fig:stateforint} has to be achieved so the interrupt \bltInlineVerb{int 0x80} causes a shell to be opened.~\cite{pixis}~\cite{proggen-rop}

@@

From all the previous steps the payload got constructed with python~\cref{code:payloadbinsh}. As seen in the example we can define all the addresses, gadgets and other parameters as variables and reuse them in the \bltInlineVerb{pack} calls, this way changing a gadget only requires one value to be changed. The output gets written into a file and can then be used for the attack.

@@

The only protections that had to be disabled were stack canaries and ASLR. It is likely that there are systems still in use today which are vulnerable to this kind of attack due to not having these protections or the protections themselves being attackable. Since it can allow arbitrary code execution it is very important to identify these devices and patch or replace them.

@@

The information wether ROP works with ASLR enabled is inconsistent. While trying this attack with \bltInlineVerb{/proc/sys/kernel/randomize_va_space} set to \bltInlineVerb{2} meaning full randomization the attack still seemed to work. The inconsistent information probably arises due to different approaches being used. With executables that have PIE enabled ROP is still possible but only with ASLR disabled~\cite{el-sherei}. With the compiler option \bltInlineVerb{-static} used for this example PIE is implicitly disabled and ASLR seems to have no effect on the exploit. This is likely because the ASLR settings \Verb+1+ and \Verb+2+ only randomize shared libraries and PIE binaries~\cite{rny}, meaning ASLR is disabled in this example, even when the value in \bltInlineVerb{/proc/sys/kernel/randomize_va_space} is \bltInlineVerb{1} or \bltInlineVerb{2}.

\section{Protection}
\paragraph{Stack canaries}
Stack canaries are one of the most effective approaches against ROP, they are enabled by default and prevent most forms of buffer overflows, however, stack canaries can be based on a small entropy pool and can therfore be bruteforced with an effort significantly smaller than regular bruteforcing. Depending on the target it can still be profitable and possible to bruteforce it even with a big entropy pool.
\paragraph{NX}
The activation of the NX bit has no effect on ROP since the program never executes code outside the segments marked with the \bltInlineVerb{CODE} flag like in a classical stack overflow attack.~\cite{ropsla}
\paragraph{ASLR}
According to a paper by Hovav Shacham et al. ASLR is a good protection against ROP in 64 bit binaries assuming no side channel leakage since 40 bit are available for randomizations of the libraries and code locations, however, 32 Bit binaries only use 16 Bit for randomization. Because of that they were able to perform a buffer overflow attack like ret2libc on an Apache server with an average of 216 seconds.~\cite{aslr}

\section{Conclusion}
As it has been demonstrated Return Oriented Programming is a powerful exploitation technique which should be taken seriously. As we were able to see the only way to make a binary relatively safe is to compile it as 64 Bit with stack canaries, bounds checking and ASLR enabled, though even then side channel attacks, bugs and bruteforcing based on a poor entropy pool may make an attack possible. Replacing or patching vulnerable devices is very important and more research on this topic paired with direct action in systems design may be necessary to keep up with black-hats.
%%%% 8. BILBIOGRAPHY %%%%
\bibliographystyle{alpha}
\bibliography{refs}
%%%% NOTES
% - Download abbrev3.bib and crypto.bib from https://cryptobib.di.ens.fr/
% - Use bilbio.bib for additional references not in the cryptobib database.
%   If possible, take them from DBLP.
\end{document}
